---
title: "Code of geneFiltering()"
subtitle: Workflow run by geneFiltering() 
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_float: false
    css: corrected.css
  html_notebook:
    toc: yes
  pdf_document:
    toc: yes
package: '`r pkg_ver("SCENIC")`'
vignette: >
  %\VignetteIndexEntry{Internal code for: geneFiltering()} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Suppress loading messages when building the HTML
suppressPackageStartupMessages({
  library(SCENIC)
  library(AUCell)
  library(RcisTarget)
  library(SingleCellExperiment)
})
```

This tutorial provides the detailed explanation of `geneFiltering()`: Filter-out genes (previous to GRN-inference) based on the counts and number of cells in which they are detected.

All the code below is the content of the function `geneFiltering()`. This tutorial is meant for advanced users, who want know the details about what this function does internally, or to modify the workflow. There is no need to follow this tutorial for a regular run of SCENIC (see `vignette("SCENIC_Running")`).


# Gene filtering

See `SCENIC_running.Rmd` vignette for the description/motivation of this function.

# Input

```{r, eval=FALSE}
setwd("SCENIC_MouseBrain")
scenicOptions <- readRDS("int/scenicOptions.Rds")

load("data/sceMouseBrain.RData")
exprMat <- counts(sceMouseBrain)
```


# geneFiltering() code:
```{r, eval=FALSE}
# Default values:
minCountsPerGene <- 3*.01*ncol(exprMat)
minSamples <- ncol(exprMat)*.01
outFile_exprMatTxt <- NULL # Filtered expression matrix is not saved

dbFilePath <- getDatabases(scenicOptions)[[1]]
outFile_genesKept <- getIntName(scenicOptions, "genesKept")
```

```{r, eval=FALSE}
# Calculate stats
nCountsPerGene <- rowSums(exprMat, na.rm = T)
nCellsPerGene <- rowSums(exprMat>0, na.rm = T)

## Show info
message("Maximum value in the expression matrix: ", max(exprMat, na.rm=T))
message("Ratio of detected vs non-detected: ", signif(sum(exprMat>0, na.rm=T) / sum(exprMat==0, na.rm=T), 2))
message("Number of counts (in the dataset units) per gene:")
print(summary(nCountsPerGene))
message("Number of cells in which each gene is detected:")
print(summary(nCellsPerGene))

## Filter
message("\nNumber of genes left after applying the following filters (sequential):")
# First filter
# minCountsPerGene <- 3*.01*ncol(exprMat)
genesLeft_minReads <- names(nCountsPerGene)[which(nCountsPerGene > minCountsPerGene)]
message("\t", length(genesLeft_minReads), "\tgenes with counts per gene > ", minCountsPerGene)

# Second filter
# minSamples <- ncol(exprMat)*.01
nCellsPerGene2 <- nCellsPerGene[genesLeft_minReads]
genesLeft_minCells <- names(nCellsPerGene2)[which(nCellsPerGene2 > minSamples)]
message("\t", length(genesLeft_minCells), "\tgenes detected in more than ",minSamples," cells")

# Exclude genes missing from database:
library(RcisTarget)
motifRankings <- importRankings(dbFilePath) # either one, they should have the same genes
genesInDatabase <- colnames(getRanking(motifRankings))

genesLeft_minCells_inDatabases <- genesLeft_minCells[which(genesLeft_minCells %in% genesInDatabase)]
message("\t", length(genesLeft_minCells_inDatabases), "\tgenes available in RcisTarget database")

genesKept <- genesLeft_minCells_inDatabases

if(is.null(outFile_genesKept)) 
  saveRDS(genesKept, file=outFile_genesKept)
```

Return value: `genesKept`

