---
title: "SCENIC step 1 (Part 2): Create co-expression modules"
abstract: "The first step of SCENIC workflow aims to infer potential transcription factor targets based on co-expression. On the first part, we ran GENIE3. Here, we will convert GENIE3 output into gene-sets with putative TF targets (co-expression modules)."
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
  html_notebook:
    toc: yes
  pdf_document:
    toc: yes
package: '`r pkg_ver("SCENIC")`'
vignette: >
  %\VignetteIndexEntry{SCENIC Step1.2: Transform GENIE3 output into co-expression modules} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Suppress loading messages when building the HTML
suppressPackageStartupMessages({
    library(Biobase)
    library(data.table)
    library(reshape2)
    library(GENIE3)
})

# Do not convert strings to factors (IMPORTANT! Specially if reading-in GENIE3 text output)
options(stringsAsFactors=FALSE)

# To build a personalized report, update this working directory:
knitr::opts_knit$set(root.dir = 'SCENIC_MouseBrain')
```

> The code from this file is included in the `runSCENIC()` function. It is not required to run it separatelly. This tutorial assumes you have read the introduction to SCENIC and run the previous steps: `vignette(package="SCENIC")`.

# Step 1 - Part2: Transform GENIE3 output into co-expression modules 

Once GENIE3 is run, it provides an *adjacency matrix* including the potential regulators for each gene and their *weight*. The weight represents the relevance that the transcription factor (regulator) has in the prediction of the target. However, this output includes all possible pairs of genes and regulators (even if the weight is very low) and there is not a specific method or clear recommendation to select a subset of them. 
The most direct way to obtain the relevant links is to keep only those with a weight over a given threshold. These links can then be split by the transcription factor, to obtain potential targets for each TF. 

We explored several ways to determine the threshold (e.g. looking at the rankings, distributions, and output after pruning with RcisTarget), but there was no specific method that was optimum for all cases. On the contrary, the best results applying SCENIC resulted from the combination of several strategies. In this way, we have finally opted for building multiple gene-sets of potential targets for each transcription factor: [**a**] setting several weight thresholds, [**b**] taking the top 50 targets for each TF, and [**c**] keeping only the top regulators for each target gene (then, split by TF). In all these cases, only the links with weight>0.001 are taken into account. 

The first method to create the TF-modules is to select the best targets for each transcription factor:

1. Targets with **weight > 0.001**

2. Targets with **weight > 0.005**

3. **Top 50 targets** (targets with highest weight)

The alternative way to create the TF-modules is to select the best regulators for each gene (this is actually how GENIE3 internally works). Then, these targets can be assigned back to each TF to form the TF-modules. In this way we will create three more gene-sets:

4. Targets for which the TF is within its **top 5 regulators**

5. Targets for which the TF is within its **top 10 regulators**

6. Targets for which the TF is within its **top 50 regulators**

## Load output from GENIE3
GENIE3 default output is in the form of weight matrix, we will convert it to a link list:
```{r loadGenie3, eval=TRUE}
library(GENIE3)
# Convert the weight matrix into links:
load("int/1.3_GENIE3_weightMatrix.RData")
linkList <- getLinkList(weightMatrix, threshold=0.001) # (slighly faster)
# linkList <- getLinkList(weightMatrix)
colnames(linkList) <- c("TF", "Target", "weight")
# order by weight
linkList <- linkList[order(linkList[,"weight"], decreasing=TRUE),]
save(linkList, file="int/1.5_GENIE3_linkList.RData")
```

The link list is the main output from GENIE3. The first column contains the TF (regulatory gene), the second column shows the target gene, and the last column indicates the weight of the link:
```{r checkLinkList}
load("int/1.5_GENIE3_linkList.RData")
dim(linkList)
head(linkList)
```

## Creating TF modules (potential TF-targets)

### Build gene-sets

To build the gene-sets (TF modules), the first step is to filter the link-list to keep only the links with weight over the *minimum* threshold (in this case `0.001`, check whether this value is also valid for your dataset). 

```{r weightStats}
quantile(linkList$weight, probs=c(0.75, 0.90))
plot(linkList$weight[1:1000000], type="l", ylim=c(0, max(linkList$weight)), main="Weight of the links",
     ylab="Weight", xlab="Links sorted decreasingly")
abline(h=0.001, col="blue") # Threshold
sum(linkList$weight>0.001)/nrow(linkList)
```

Keep only the links over `0.001`:
```{r filterLinkList}
linkList_001 <- linkList[which(linkList[,"weight"]>0.001),]
# Number of links over the threshold: 
nrow(linkList_001) 
```

Create the gene-sets & save:
```{r splitLinkList, eval=TRUE}
tfModules <- list()

linkList_001$TF <- as.character(linkList_001$TF)
linkList_001$Target <- as.character(linkList_001$Target)
  
#### Create TF-modules:
# 1: Weight > 0.001 (filtered in previous step) 
tfModules[["w001"]] <- split(linkList_001$Target, factor(linkList_001$TF))

# 2: Weight > 0.005
llminW <- linkList_001[which(linkList_001[,"weight"]>0.005),]
tfModules[["w005"]] <- split(llminW$Target, factor(llminW$TF))

# 3: Top 50 targets for each TF
# ("w001" should be ordered decreasingly by weight)
tfModules[["top50"]] <- lapply(tfModules[["w001"]], function(x) x[1:(min(length(x), 50))])

# 4-6: Top regulators per target 
# (linkList_001 should be ordered by weight!)
linkList_001_byTarget <- split(linkList_001, factor(linkList_001$Target))
save(linkList_001_byTarget, file="int/1.5_linkList_001_byTarget.RData")

nTopTfs <- c(5, 10, 50)
nTopTfs <- setNames(nTopTfs, paste("top", nTopTfs, "perTarget", sep=""))

library(reshape2); library(data.table)
topTFsperTarget <- lapply(linkList_001_byTarget, function(llt) {
   nTFs <- nTopTfs[which(nTopTfs <= nrow(llt))]
   melt(lapply(nTFs, function(x) llt[1:x,"TF"]))
})
topTFsperTarget <- topTFsperTarget[which(!sapply(sapply(topTFsperTarget, nrow), is.null))]
topTFsperTarget.asDf <-  data.frame(rbindlist(topTFsperTarget, idcol=TRUE))
head(topTFsperTarget.asDf)
colnames(topTFsperTarget.asDf) <- c("Target", "TF", "method")

# Merge the all the gene-sets:
tfModules.melted <- melt(tfModules)
colnames(tfModules.melted) <- c("Target", "TF", "method")
tfModules <- rbind(tfModules.melted, topTFsperTarget.asDf)

save(tfModules, file="int/1.6_tfModules.RData")
```

```{r showTfModules}
load("int/1.6_tfModules.RData")
# Basic counts:
rbind(nGeneSets=nrow(tfModules), 
      nTFs=length(unique(tfModules$TF)), 
      nTargets=length(unique(tfModules$Target)))
```

### Split into positive- and negative-correlated targets

The TF-modules from the previous step can already be analyzed for motif enrichment (Step 2 of SCENIC). However, GENIE3 can detect both positive and negative associations. In order to distinguish potential activation from repression, we split the targets into positive- and negative- correlated targets.

#### Split TF targets according to correlation
To easily split the TF targets into positive- or negative-correlated, we will add this information into the TF-modules table. The new table will contain a column (`$corr`) with value `1` if the correlation between the TF and the target is > 0.03, `-1` if the correlation is < -0.03 and `0` otherwise.

```{r addCorr, eval=TRUE}
load("int/1.4_corrMat.RData")
# Keep only correlation between TFs and potential targets
tfs <- unique(tfModules$TF)
corrMat <- corrMat[tfs,]

# Split TF modules according to correlation
tfModules_byTF <- split(tfModules, factor(tfModules$TF))
tfModules_withCorr_byTF <- lapply(tfModules_byTF, function(tfGeneSets)
{
    tf <- unique(tfGeneSets$TF)
    targets <- tfGeneSets$Target
    cbind(tfGeneSets, corr=c(as.numeric(corrMat[tf,targets] > 0.03) - as.numeric(corrMat[tf,targets] < -0.03)))
})
tfModules_withCorr <- data.frame(rbindlist(tfModules_withCorr_byTF))
save(tfModules_withCorr, file="int/1.7_tfModules_withCorr.RData")
```

```{r showTfModules_withCorr}
load("int/1.7_tfModules_withCorr.RData")
head(tfModules_withCorr)
dim(tfModules_withCorr)
```

From this table, we can easily select gene-sets associated to each TF (i.e. `split(tfModules_withCorr$Target, tfModules_withCorr$TF)`).
In this way, we obtain a list of potential targets for each TF based on their co-expression (*TF co-expression modules*). 

In the next step, we will use RcisTarget to check which of these targets present enrichment of the motifs of the corresponding TF.

# SessionInfo
```{r sessionInfo}
date()
sessionInfo()
```
